---
title: 'Basic graphics with tidy data, or: why starting tidy makes life in R palatable'
author: "Elizabeth Wickes, wickes1@illinois.edu, @elliewix"
output:
  pdf_document:
    fig_height: 3
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 6
---

## Spreadsheet activity

Handout for the "Creating Tidy Data" workshop of the 2016 Midwest Data Librarian Symposium (MDLS). 
GitHub Repository: https://github.com/elliewix/but-i-just-want-to-make-a-graph

Open up the Excel file included within the GitHub repository.  Take a few moments to explore it, the data included, structure, etc.  Now, try to think about how you could reorganize it into a single data file.  Work with 1-2 people sitting next to you.  Determine what the new headers should be and write them down below.

&nbsp;

&nbsp;

&nbsp;

Create a new tab and begin recreating the new data file if you have time remaining.

## R graphing activity

You should follow along in RStudio with these commands. You will be using a data file called `cleanedsurvey.csv` located within the GitHub repository.

### Load the library

You should have installed this with `install.packages("ggplot2")`.  You call `library()` with the module name, without quotes, to load the library.

```{r, warning=FALSE}
library(ggplot2)
```


### Load the data

Use Tools -> Import Dataset -> From text file...

Select the files, and accept all the default options in the wizard.  You can play around with these options and view the changes in the preview box.

```{r, warning=FALSE, error=FALSE}
cleanedsurvey <- read.csv("cleanedsurvey.csv")
```

This creates an object in memory, called `cleanedsurvey`, that represents the data.frame containing the data.

### Make a basic histogram

This will be using ggplot2 to create a histogram.  They usually start with a `ggplot()` statement that declares where the data frame is, then you use `+` to add graphic elements to the plot. You can layer these elements as desired.

You can see in the code below that we are declaring `data = cleanedsurvey` in `ggplot()`, but not selecting any of the data columns yet.  Just providing the data frame.

`+` then adds a `geom_histogram()` graphical layer, which is a histogram.  We tell it that we want `x = Weight`. `Weight` is the column name without quotes.  The case must also match what appears in the data.frame it is contained within. There are many types of graphical elements (http://docs.ggplot2.org/current/) within the ggplot2 library.

You'll see below that the plot is giving us a message that it is using a default bin value.

```{r, warning=FALSE, error=FALSE, results='hide'}
ggplot(data = cleanedsurvey) + geom_histogram(aes(x = Weight))
```

As that error message hinted, `geom_histogram()` has additional options that you can play with.  Place your curser just inside the last set of `()` press tab to see them.  Let's play with binwidth.

```{r, warning=FALSE, error=FALSE}
ggplot(data = cleanedsurvey) + geom_histogram(aes(x = Weight), binwidth = 2)
```

### Now you

* Change some of the binwidth values to find the right balance for the data you are looking at
* Change the `x =` value to other columns and see what works and what doesn't.

* Which columns worked?  Which columns didn't?
* What sort of errors did you get when you tried a column that didn't work?
* What other sort of errors did you see in this process?

### Scatterplots and grouping

We can also use scatterplots to look at numerical data.

You can see that the syntax is already changing on us, but at least the x and y labels are showing our nice column headers.

```{r, warning=FALSE, error=FALSE, results='hide'}

ggplot(cleanedsurvey, aes(x = Weight, y = Plot)) + geom_point()

```

This is nice, but hard to tell a bit of what's going on.  Let's group this in a few ways.  

Because our code is getting long, we can save our basic ggplot code as a variable and reference it.  We're going to call our ggplot() object `p`.  Nothing will happen when we first create `p` because we must call p to make it appear.  Calling `p` alone will show a blank graph. We need to add our graphical element layer to see the actual chart.

```{r, warning=FALSE, error=FALSE, results='hide'}

p = ggplot(cleanedsurvey, aes(x = Weight, y = Plot)) # nothing will be plotted yet

p + geom_point() # the plot will appear when you evaluate it with a geometry element

```

Now we're seeing the power of having our data already in tidy format.  Because all our labels are in a single column, associated with each observation, we can just point R to that column and have nice labels and colors without too much work.

```{r, warning=FALSE, error=FALSE, results='hide'}

p + geom_point(aes(color = Species)) # color by the Species column
#benign change

```

### Now you

* Try changing the color variable to other columns within the data file.
* What worked? What didn't?
* What were the errors that you encountered?

